---
title: "Estimating Parameters with IRTpp"
author: "SICS Research Group"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Estimating Parameters with IRTpp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Estimating Item Parameters according to a IRT model with the IRTpp package
========================================================



The main purpose of the irtpp package is to estimate item parameters according to a IRT model, 
to begin the paremeter estimation for a test, use the function `irtpp`

```{r}
library(IRTpp)
tst = simulateTest(model="3PL")

## Calibrate this test with a 2PL model.
est = irtpp(tst$test,"1PL")
```

Now `est` holds the return of the estimation procedure 

```{r}
names(est)
est$z
est$LL
```

in `est$z` there is a named list of the IRT parameters calibrated for this test. It is also easy to visualize all the Item Caracteristic Curves (ICC) of the test, with the function `test.plot`, this is a visual resume of the parameters.

```{r}
test.plot(est$z)
```

Alternatively , `irtpp` offers other options such as displaying loglikelihood based statistics , reading from files or changing the initial values of the estimation procedure (for advanced uses).

```{r}
## Calibrating the same test under a 3PL model and displaying the AIC and BIC statistics
est = irtpp(tst$test,"3PL", loglikflag=T)
```

The item caracteristic curve of all the items of the test can be drawn to compare between IRT models.

```{r}
test.plot(est$z)
```

Or for an specific item: 

```{r}
test.plot(est$z,2)
```

## Estimating latent traits of the individuals

IRTpp package also supports estimating latent traits of individuals according to IRT methodologies. User the function `individual.traits` to estimate the individual latent traits for a test.

```{r}
zz = parameter.matrix(est$z,byrow = F)
th = individual.traits(model="3PL", itempars = zz,method = "EAP",dataset = tst$test, probability_matrix = est$prob_mat)

##The latent traits.
hist(th[,ncol(th)],breaks=40)
```
